msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-14 paper.tex

The more recent trend has been to move away from this file and command-line
driven approach and to instead provide direct interfaces to the simulation
engines via an Application Programming Interface (API)~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}. This
approach greatly increases the flexibility of simulations, and allows all
processing for many simulation analyses to be performed in a single script. The
primary interface for \msprime\ is through a well documented and stable Python
API, which has allowed a number of downstream tools to be built on
top of
% Found these through the Dependent packages on GitHub
it~\citep{kamm2020efficiently,adrion2020predicting,adrion2020community,
rivera2021simulation,de2021geonomics}
as well as numerous easily reproducible simulation
% TODO: more papers with exemplorary reproducible simulation pipelines based
% on msprime!
pipelines~\citep[e.g.][]{martin2017human}.
We also provide an \ms\ compatible
command line interface to support existing workflows, as well as the
\texttt{msp simulate} command that supports specifying demographic models in
Demes format~\citep{gower2021demes}.

A major change for the 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
resolving a major source of confusion and error. The 0.x APIs are fully
supported and tested, and will be maintained for the foreseeable future.

>>>>

The more recent trend has been to move away from this file and command-line
driven approach and to instead provide direct interfaces to the simulation
engines via an Application Programming Interface (API)~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
The primary interface for \msprime\ is through a thoroughly documented and
stable Python
API, which has allowed a number of downstream tools to be built on top of
% Found these through the Dependent packages on GitHub
it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
adrion2020community,adrion2020predicting, kamm2020efficiently,
mckenzie2020ipcoal, montinaro2020revisiting,
de2021geonomics,rivera2021simulation}.
As well as providing a stable and efficient platform for building
downstream applications, \msprime's Python API makes it much easier to
build reproducible simulation pipelines, as the entire workflow can
be encapsulated in a single script, and package and version
dependencies can be explicitly stated using the \texttt{pip}
or \texttt{conda} package managers.
For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
in the influential simulation analysis of
\cite{martin2017human} were only detected because the pipeline
could be easily run and reanalysed.

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
resolving a major source of confusion and error. The 0.x APIs are fully
supported and tested, and will be maintained for the foreseeable future.
We also provide an \ms\ compatible
command line interface to support existing workflows, and the
\texttt{msp} program that supports specifying demographic models in
Demes format~\citep{gower2021demes} from the command line.
